Reasoning About a Machine with Local Capabilities

### 2019-02-06 toplas/paper.tex
Update TOPLAS LR section with new read cond description

%subsection introduction
% \dominique{I would try to motivate the LR defs by saying that the LR is
%   constructed to express the FTLR and the FTLR says that if a piece of code only
%   has access to values respecting the invariants in a world, then it will
%   respect them itself too.} 
The logical relation defines semantically when values, program
counters, and configurations are capability safe. The definition is
found in \figurename{}s~\ref{fig:logrel} and~\ref{fig:perm-conds} and
we provide some explanations in the following paragraphs. For space
reasons, we omit some definitions and explain them only verbally, but
precise definitions can be found in the appendix.

First, the \emph{observation relation} $\observations$ defines what
configurations we consider safe. A configuration is safe
with respect to a world, when the execution of said configuration does not break
the memory protocols of the world. Roughly speaking, this means that when the
execution of a configuration halts, then there is a private future world that
the resulting memory satisfies. Notice that failing is considered safe behavior.
In fact, the machine often resorts to failing when an unauthorized access is
attempted, such as loading from a capability without read permission. This is
similar to \citet{Devriese:2016ObjCap}'s logical relation for an untyped
language, but unlike typical logical relations for typed languages, which
require that programs do not fail.

The \emph{register-file relation} $\stdrr$ defines safe register-files as those
that contain safe words (i.e.\ words in the $\stdvr$-relation defined below) in all registers but $\pcreg$.
The \emph{expression relation} $\stder$ defines that a word is safe to use as a
program counter if it can be plugged into a safe register file (i.e.\ a register
file in $\stdrr$) and paired with a memory satisfying the world to become a safe
configuration. Note that integers and non-executable capabilities (e.g.
$\readonly$ and $\entry$ capabilities) are considered safe program counters
because when they are plugged into a register file and paired with a memory, the
execution will immediately fail, which is safe.

The \emph{value relation} $\stdvr$ defines when words are safe.
\itoplassug{Explain why we use $\UPred{}$ with reference to the introduction of this section.}
We make the
value relation as liberal as possible by considering what is the most we can
allow an adversary to use a capability for without breaking the memory
protocols. Non-capability data is always safe because it provides no
authority. Capabilities give the authority to manipulate memory and potentially
break memory protocols, so they need to satisfy certain conditions to be
safe. In \figurename~\ref{fig:perm-conds}, we define such a condition for each
kind of permission a capability can have.

For capabilities with read permission, the $\readCond{}$ ensures that it can
only be used to read safe words, i.e.\ words in the value relation. To guarantee
this, we require that the addressed memory is governed by a region $W(r)$ that
imposes safety as a requirement on the values contained. This safety requirement
is formulated in terms of a standard region $\iota^\pwl_{[\start,\addrend]}$. It simply
requires all the words in the range $[\start,\addrend]$ to be safe, i.e.\ in the
value relation. Requiring that $W(r)\nsubsim[n]\iota^\pwl_{[\start,\addrend]}$
means that $W(r)$ must accept only memory segments containing safe values like
$\iota^\pwl_{[\start,\addrend]}$, but can be even more restrictive if needed
(the $\nsubsim[n]$-relation is defined in Appendix~\ref{app:logical-relation}). The
read condition also takes into account the locality of the capability because,
generally speaking, global capabilities should only depend on permanent regions.
Concretely, we use the function $\var{localityReg}(\gl,W)$, which projects out
all active (non-revoked) regions when the locality $\gl$ is local, but only the
permanent regions when $\gl$ is global. The definition of the standard region
$\iota^\pwl_{[\start,\addrend]}$ can be found
in \figurename~\ref{fig:std-reg}; it makes use of the isomorphism $\xi$ from
Theorem~\ref{thm:world-existence}.

>>>>

%subsection introduction
% \dominique{I would try to motivate the LR defs by saying that the LR is
%   constructed to express the FTLR and the FTLR says that if a piece of code only
%   has access to values respecting the invariants in a world, then it will
%   respect them itself too.} 
The logical relation defines semantically when values, program
counters, and configurations are capability safe. The logical relation is defined
found in \figurename{}s~\ref{fig:logrel} and~\ref{fig:perm-conds}, and
we provide some explanations in the following paragraphs. For space
reasons, we omit some definitions and explain them only verbally, but
precise definitions can be found in the appendix.

First, the \emph{observation relation} $\observations$ defines what
configurations we consider safe. A configuration is safe
with respect to a world, when the execution of said configuration does not break
the memory protocols of the world. Roughly speaking, this means that when the
execution of a configuration halts, then there is a private future world that
the resulting memory satisfies. Notice that failing is considered safe behavior.
In fact, the machine often resorts to failing when an unauthorized access is
attempted such as loading from a capability without read permission. This is
similar to \citet{Devriese:2016ObjCap}'s logical relation for an untyped
language but unlike typical logical relations for typed languages that
require programs to not fail.

The \emph{register-file relation} $\stdrr$ defines safe register-files as those that contain safe words (i.e.\ words in the $\stdvr$-relation defined below) in all registers but the $\pcreg$-register.
The \emph{expression relation} $\stder$ defines what words are safe to use as a program counter.
A word is safe to use as a program counter when it can be used to form a safe configuration by plugging it into the $\pcreg$-register of a safe register file (i.e.\ a register file in $\stdrr$) and pairing it with a memory satisfying the world.
Note that integers and non-executable capabilities (e.g.\ $\readonly$ and $\entry$ capabilities) are considered safe program counters because when they are plugged into a register file and paired with a memory, the execution will immediately fail, which is safe.

The \emph{value relation} $\stdvr$ defines when words are safe.
\itoplassug{Explain why we use $\UPred{}$ with reference to the introduction of this section.}
We make the value relation as liberal as possible by defining it based on the principle ``what is the most we can allow an adversary to use a capability for without breaking the memory protocols.''

Non-capability data is always safe because it provides no authority.
Capabilities give the authority to manipulate memory and potentially break memory protocols, so they need to satisfy certain conditions to be safe.
In \figurename~\ref{fig:perm-conds}, we define such a condition for each kind of permission a capability can have.

Capabilities with read permission cannot directly break memory protocols because they cannot make changes to the memory.
However, a read capability could be used to read a capability that can break memory protocols.
For this reason, a capability with read permissions is only safe when it can be used only to read safe capabilities.
This is captured by the $\readCond{}$-condition. 
More precisely, the condition requires the memory addressed by the capability to actually be governed by some region $W(r)$.
This region should at least require all the memories accepted by the state interpretation function to contain safe words but stricter requirements may be imposed on the memories.
This is expressed in terms of an upper bound on the possibly permitted memories defined in terms of a so called ``standard region''.
In this case, the standard region is $\iota^\pwl_{[\start,\addrend]}$ (defined in \figurename~\ref{fig:std-reg}). 
The standard region only accept memories with address range $[\start,\addrend]$.
More importantly, the standard region's state interpretation function only accepts memory segments that only contain safe words.
The relation the actual region $W(r)$ and the upper bound $\iota^\pwl_{[\start,\addrend]}$ is $\iota^\pwl_{[\start,\addrend]}$ (defined in Appendix~\ref{app:logical-relation}).
Essentially, the relation requires the two regions state transition systems to be the same. Further in the current state of region $W(r)$ and in any world, the state interpretation function should only allow memories that are also allowed in the standard region.

The read condition also takes the locality of a read capability into account because, generally speaking, generally speaking, global capabilities should only depend on permanent regions.
In $\readCond{}$, this is manifested by $\var{localityReg}(\gl,W)$ which projects out all active (non-revoked) regions when the locality $\gl$ is local, but only the permanent regions when $\gl$ is global.
